---
title: "Bayesian CFA"
format:
  html:
    theme: default
---

```{r message = FALSE}
library(tidyverse)
library(lavaan)
library(brms)
library(tidybayes)

# brms backend specifications
options(brms.backend = "cmdstanr")
options(mc.cores = parallel::detectCores())
```

Can brms replicate results from lavaan?

{{brms}} doesn't officially have SEM capabilities [(but they do seem to be coming soon!)](https://github.com/paul-buerkner/brms/issues/304). But [STAN forum contributer Jack Bailey](https://discourse.mc-stan.org/t/confirmatory-factor-analysis-using-brms/23139) has 'patanté' a solution using artful prior specifications.

To demonstrate, we can simulate some data using {{lavaan}} and show that `brms::brm()` is able to recover the 'true' factor loadings.

## First Example: Simple Factor Structure

First we can simulate some data with a simple factor structure using lavaan's handy `simulateData()` function:
```{r}

pop.model <- ' 
  f1 =~ .8*m1 + .1*m2 + .6*m3 + .2*m4 + .9*m5 + -.4*m6
'

# Simulate data from the model
fake_dat <- lavaan::simulateData(model = pop.model, sample.nobs = 4000)

# Visualize the measured dat
fake_dat |>

  select(m1, m2, m3, m4, m5, m6) |>
  
  pivot_longer(everything(), names_to = "var", values_to = "measurement") |>

  ggplot() +
  geom_histogram(aes(x = measurement)) +
  facet_wrap(~var) +
  theme_bw()

```

Unsurprisingly, the lavaan is able to recover the true parameters for the data it generated.

```{r}

# Fit the model
fit_1 <- cfa(pop.model, data = fake_dat)

# Make sure it works
fit_1 |> broom::tidy()

```

But can brms?

```{r eval = FALSE}

# Add a latent variable to the dataset
fake_dat$f1 <- NA_real_

bfit.1 <- brm(
    formula =
      bf(m1 ~ 0 + mi(f1)) +
      bf(m2 ~ 0 + mi(f1)) +
      bf(m3 ~ 0 + mi(f1)) +
      bf(m4 ~ 0 + mi(f1)) +
      bf(m5 ~ 0 + mi(f1)) +
      bf(m6 ~ 0 + mi(f1)) +
      bf(f1| mi() ~ 1) + 
      set_rescor(rescor = FALSE),
    family = gaussian(),
    prior =
      prior(constant(1), class = "b", resp = "m1") +
      prior(constant(1), class = "sigma", resp = "m1") +
      prior(normal(0, 10), class = "b", resp = "m2") +
      prior(constant(1), class = "sigma", resp = "m2") +
      prior(normal(0, 10), class = "b", resp = "m3") +
      prior(constant(1), class = "sigma", resp = "m3") +
      prior(normal(0, 10), class = "b", resp = "m4") +
      prior(constant(1), class = "sigma", resp = "m4") +
      prior(normal(0, 10), class = "b", resp = "m5") +
      prior(constant(1), class = "sigma", resp = "m5") +
      prior(normal(0, 10), class = "b", resp = "m6") +
      prior(constant(1), class = "sigma", resp = "m6") +
      prior(normal(0, 10), class = "Intercept", resp = "f1") +
      prior(cauchy(0, 1), class = "sigma", resp = "f1"),
    data = fake_dat,
    warmup = 1000,
    iter = 6000,
    file = "source/fits/b07.01.rds"
  )

```

A Bayesian missing data model treats each missing observation as a parameter. And since the factor is missing in each row of the dataset, this means we end up with 6000 samples for each row of data we have, resulting in a pretty big model file. This is too big for Github, so I'll only push the draws we need to replicate the figure below.

First we can put the draws we need into a tidy format and do some cleaning:

```{r}

bfit.1.samples <- bfit.1 |>

  # Get the raw MCMC samples
  gather_draws(
    bsp_m2_mif1,
    bsp_m3_mif1,
    bsp_m4_mif1,
    bsp_m5_mif1,
    bsp_m6_mif1
  ) |>

  # Do some renaming for clarity
  mutate(.variable = case_when(
    .variable == "bsp_m2_mif1" ~ "Loading for m2",
    .variable == "bsp_m3_mif1" ~ "Loading for m3",
    .variable == "bsp_m4_mif1" ~ "Loading for m4",
    .variable == "bsp_m5_mif1" ~ "Loading for m5",
    .variable == "bsp_m6_mif1" ~ "Loading for m6"
  )) |> 

  # Add the true factor loadings from above
  mutate(true_loading = case_when(
    .variable == "Loading for m2" ~ .1,
    .variable == "Loading for m3" ~ .6,
    .variable == "Loading for m4" ~ .2,
    .variable == "Loading for m5" ~ .9,
    .variable == "Loading for m6" ~ -.4
  )) 

# Save the tidy draws for reproducibility
saveRDS(bfit.1.samples, "source/fits/b07.01.samples.rds")
```

Now we can plot. Looks like STAN does an OK job at recovering the true factor loadings (we exclude the first loading since its loading was held constant at 1 to provide the scale for the latent variable):
```{r}
#| fig-cap: "Posterior distributions of estimated factor loadings for measured variables 2-6. The dashed line is the 'true' simulated loading for each variable."
#| fig-alt: "Posterior distributions of estimated factor loadings for measured variables 2-6. The dashed line is the 'true' simulated loading for each variable.

# Load the tidy samples
bfit.1.samples <- readRDS("source/fits/b07.01.samples.rds")

# Plot
bfit.1.samples|>

  ggplot(aes(x = .value, fill = .variable)) +
    stat_halfeye(fill = "mediumorchid") +
    geom_vline(aes(xintercept = true_loading), linetype = 2) + 
    scale_x_continuous(expand = c(0, 0.015)) +
    scale_y_continuous(expand = c(0, 0.015)) +
    guides(fill = "none") +
    labs(x = "lab",
       y = NULL)  +
    facet_wrap(~.variable) +
    theme_minimal() + 
    theme(panel.grid.major = element_blank())
   

  

  


```

```{r}

fit3 <- brm(formula = rating ~ treat + period,
data = inhaler, family = cumulative)

fit3 |> summary()
```
## Second Example: Correlated factors

## Third Example: MTMM 